STAT 672 Final Project: Stochastic Gradient Descent

### 2018-04-15 paper/paper.tex
initial
rover 2.0

\DeclareMathOperator*{\argmax}{argmax}

@@

Assuming that the outcome variable follows a
particular statistical distribution aids the computation of optimal coefficients. 
For example, assumptions in ordinary least squares (OLS)
regression---assumptions that readers almost certainly are familiar with and so will not be
repeated here---allow a closed form solution. Suppose we have $n$
observations. Each observation consists of an outcome variable $y_i$ and some
explanatory variables $x_{i1}, x_{i2}\ldots x_{iD}$. We thus have a vector of
outcomes $\bm{Y}_{n \times 1}$ and a feature matrix $\bm{X}_{n \times D}$. All
variables are real numbers: $\bm{X} \in \mathbb{R}^{n \times D}$, $\bm{Y} \in
\mathbb{R}^{n \times 1}$. The goal
is estimate weights $w_1 \ldots w_j \ldots w_D$, or $\bm{w}_{1 \times D}$, that relate the explanatory variables
to the outcome.\footnote{Readers may be more familiar with $\bm{\beta}$ as the
notation for coefficients. $\bm{w}$ is used here to ensure common notation with
other classes of models.} The normal equations give:

$$
\hat{\bm{w}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}
$$

Even if a parametric model does not have a closed-form solution, the parametric
assumption allows some useful optimization techniques. Consider logistic
regression. The maximum likelihood estimator (MLE) approach leads to
a system of $D$ equations. This system of
equations typically is numerically solved using the iterative Newton-Raphson
algorithm:

$$
\hat{\bm{w}}_{n+1} = \hat{\bm{w}}_{n} -
\bm{H}^{-1}(\hat{\bm{w}}_n)\bm{J}(\hat{\bm{w}}_n)
$$

$\bm{J}$ is the Jacobian (the
first derivative of the log-likelihood function $l$ with respect to each $w_j$)
and $\bm{H}$ is the Hessian (the second derivative of $l$ with respect to $w_j,
w_{j'}$). The practicality of Newton-Raphson thus depends on whether it is convenient to
find $\bm{J}$ and $\bm{H}$. It is convenient for logistic regression
because parametric and independent-and-identically-distributed (IID) assumptions
mean $l$ is a simple sum of the log probability distribution
function (PDF, in this case binomial) for each observation. We ``know'' (assume)
the form of this PDF and so are
confident that the second derivative exists and is not too onerous to calculate. It may not be true, and hence
Newton-Raphson will not be practical, if the Jacobian or Hessian of the function we are trying to
maximize or minimize (perhaps not $l$) are non-existent or cumbersome.

The need to conduct optimization in non-parametric settings is a chief
motivation for gradient descent (GD), of which SGD is a variant. In
non-parametric settings---most notably supervised and unsupervised statistical
learning, in which we again seek to find optimal $\hat{\bm{w}}$ to relate input
variables to output variables for the purposes of classification or regression---there 
typically is no closed form solution for $\hat{\bm{w}}$. It
also may not be convenient to find and evaluate the Hessian, making
Newton-Raphson undesirable. SGD does not require any parametric assumptions. 
In its most basic form, SGD only requires finding the gradient (though some extensions
do need the Hessian or an approximation to it). 
SGD thus is well-suited for non-parametric settings.

\subsubsection{Computational complexity}

How an optimization technique scales with $n$ and $D$ is another important
consideration. It is little comfort if a method reaches the correct solution but
requires an excessive amount of time to do so. ``Plain'' or
``batch'' GD requires evaluating the gradient for every single observation,
every single iteration, until the algorithm converges. For example, for a
dataset of $n=10^6$ that required 25 iterations to converge, batch GD would require 
evaluating the gradient $25 \times 10^6$ times. This scaling with
$n$ can cause untenably long computation time.

@@

\subsection{Setup and Notation}

Consider a typical soft-margin support vector machine (SVM) problem.

\subsection{Basic Form}

\cite{bottou2010large}
\cite{bottou2012stochastic}
\cite{boyd2004convex}
\cite{dal2015calibrating}

@@

@@

The basic SGD algorithm has been extended in different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
development. This sub-section covers some of the more interesting developments.

\subsubsection{Step Size (Learning Rate)}

@@

@@

Consider the SVM problem posed in Section 2.

>>>>

\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}

@@

Assuming that the outcome variable follows a
particular statistical distribution aids the computation of optimal coefficients. 
For example, assumptions in ordinary least squares (OLS)
regression---assumptions that readers almost certainly are familiar with and so will not be
repeated here---allow a closed form solution. The optimization problem of
choosing coefficient $\hat{\bm{\beta}}$ that minimize squared error is solved
by $\hat{\bm{\beta}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ (where $\bm{X}$ is the
feature matrix and $\bm{Y}$ the outcome variable).

Even if a parametric model does not have a closed-form solution, the parametric
assumption allows some useful optimization techniques. Consider logistic
regression. The maximum likelihood estimator (MLE) approach for estimating
coefficients leads to a system of $D$ equations. This system of
equations typically is numerically solved using the iterative Newton-Raphson
algorithm:

$$
\hat{\bm{\beta}}_{n+1} = \hat{\bm{\beta}}_{n} -
\bm{H}^{-1}(\hat{\bm{\beta}}_n)\bm{J}(\hat{\bm{\beta}}_n)
$$

$\bm{J}$ is the Jacobian (the
first derivative of the log-likelihood function $l$ with respect to each $w_j$)
and $\bm{H}$ is the Hessian (the second derivative of $l$ with respect to $w_j,
w_{j'}$). The practicality of Newton-Raphson thus depends on whether it is convenient to
find $\bm{J}$ and $\bm{H}$. It is convenient for logistic regression
because parametric and independent-and-identically-distributed (IID) assumptions
mean $l$ is a simple sum of the log probability distribution
function (PDF, in this case binomial) for each observation. We ``know'' (assume)
the form of this PDF and so are
confident that the second derivative exists and is not too onerous to calculate.
In non-parametric settings, we often cannot be so certain and face the possibility of
$\bm{H}$ being non-existent or cumbersome.

The need to conduct optimization in non-parametric settings is a chief
motivation for gradient descent (GD), of which SGD is a variant. In
non-parametric settings---most notably supervised and unsupervised statistical
learning, in which we again seek to find optimal coefficients to relate input
variables to output variables for the purposes of classification or regression---there 
typically is no closed form solution for the coefficients. It
also may not be convenient to find and evaluate the Hessian, making
Newton-Raphson undesirable. SGD does not require any parametric assumptions. 
In its most basic form, SGD only requires finding the gradient (though some extensions
do need the Hessian or an approximation to it). 
SGD thus is well-suited for non-parametric settings.

\subsubsection{Computational Complexity}

How an optimization technique scales with sample size $n$ is another important
consideration. It is little comfort if a method reaches the correct solution but
requires an excessive amount of time to do so. ``Plain'' or
``batch'' GD requires evaluating the gradient for every single observation,
every single iteration, until the algorithm converges. For example, for a
dataset of $n=10^6$ that required 25 iterations to converge, batch GD would require 
evaluating the gradient $25 \times 10^6$ times. This scaling with
$n$ can cause untenably long computation time.

@@

\subsection{Basic setup}

Loosely following the set-up of \cite{bottou2010large}, consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
We have a hypothesis class $\mathcal{F}$ consisting of
various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
expresses the cost of mis-classification. Assume for now that $L$ is convex. 
We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

\begin{equation}
	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
\end{equation}

Because $L$ is convex, basic calculus tells us that there is one critical point,
it is the global minimum, and it is located where the gradient of the loss
function with respect to $\bm{w}$, $\nabla_{\bm{w}}$, is zero. GD is an iterative
algorithm to numerically approximate this point:

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

$t$ refers to iterations. $\gamma$ is a parameter controlling the step
size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. 
The algorithm stops when the gradient is
arbitarily close to zero; or, put differently, when the difference between this
iteration's estimate and the last iteration's estimate is arbitrarily small.

\begin{equation}
	\bm{w}_t - \bm{w}_{t-1} \leq \epsilon
\end{equation}

Note that the batch GD algorithm presented in (2) requires evaluating the
gradient for every single observation $i$. In a large-$n$ dataset, this can be
computationally infeasible. SGD's innovation is to instead evaluate only a
single randomly chosen observation $i$ at each iteration $t$, greatly lightening
the computational load.

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma 
	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

Having outlined the basic form of SGD, we now examine some of its key
properties.

@@

\subsubsection{Correctness}

\subsubsection{Speed}

@@

The basic SGD algorithm has been extended in different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
development. This sub-section covers some of the more interesting extensions.

\cite{bottou2012stochastic}
\cite{boyd2004convex}
\cite{dal2015calibrating}
\subsubsection{Step Size (Learning Rate)}

@@

\cite{polyak1992acceleration}
\cite{nesterov}

@@

The parallelization technique essentially is averaging. In line with previous
notation, suppose we have fixed learning rate $\gamma$,

Advantages
